An efficient contact value approach for the pair distribution function of the inhomogeneous hard-sphere fluid

### 2014-06-27 papers/pair-correlation/paper.tex

\begin{document}
\title{The contact value approach for the pair distribution function of the inhomogeneous hard-sphere
  fluid}

@@

\fixme{replace approximation with approach}\jeffsays{I'm stuck with this.}

@@

\section{Contact value approach}
In the approaches for the pair distribution function mentioned above,
the radial distribution function used in the approximation
was dependent upon the density averaged over some volume.  We seek to
achieve greater accuracy by making use of a function dependent upon
our averaged $g_{\sigma}(\rr)$ discussed above, which holds more
information about an inhomogeneous system than does a simple
convolution of the density.
%
We construct the CVA with the average of two radial distribution
functions, evaluated at the distance between the two points, that are
themselves functions of the averaged pair distribution function at contact
$g_{\sigma}(\rr)$ evaluated at the two points:
%
%% We therefore seek a function form
%% for our radial distribution that is dependent upon $g_{\sigma}(\rr)$
%% and that will approach the correct solution in the homogeneous limit
%% and also in the limit of contact distance between the two points.
\begin{align}
  g^{(2)}(\rr_1,\rr_2) = \frac{g(r_{12}, g_\sigma(\rr_1)) +
    g(r_{12}, g_\sigma(\rr_2))}{2}. \label{eq:g2-our-mean}
\end{align}
This contact value approach for $g^{(2)}(\rr_1,\rr_2)$ is
constructed to reproduce the exact value for the integral:
\begin{align}
  F_1^{\text{contact}} &= \tfrac12 \iint
  g^{(2)}_{HS}(\rr_1,\rr_2)n(\rr_1)n(\rr_2)\delta(|\rr_1-\rr_2|-\sigma)
  d\rr_1d\rr_2
  \label{eq:mean-field-contact}
\end{align}
which is the mean-field correction to the free energy (see
Eq.~\ref{eq:mean-field}) for a purely contact interaction.

@@

\section{Separable radial distribution function for the CVA-S}

@@

The key insight of the CVA-S is that we can implement
Eq.~\ref{eq:mean-field} on $\mathcal{O}(N\log N)$ time using Fast
Fourier Transforms (FFTs) where $N$ is proportional to the volume of
the system.  We begin by examining the CVA.  The two terms that are
averaged in Eq.~\ref{eq:CVA} give equal contributions to the integral:
\begin{align}
  F_1^{\textit{CVA}} &= \tfrac12 \!\! \iint \!\!
  g(r_{12},g_\sigma(\rr_2))n(\rr_1)n(\rr_2)\Phi(|\rr_1-\rr_2|)
  d\rr_1d\rr_2
\end{align}
When we introduce the separable form for $g(r_{12},g_\sigma)$ we can
further simplify this integral as
\begin{multline}
  F_1^{\textit{CVA-S}} =
  \sum_i \tfrac12 \!\! \iint \!\! \bigg( n(\rr_1)  \bigg) \\
                             \bigg( a_i(r_{12})\Phi(r_{12}) \bigg) \\
                             \bigg( b_i(g_\sigma(\rr_2))n(\rr_2) \bigg)
  d\rr_1d\rr_2
\end{multline}
where the functional is written as a summation of integrals, each of
which is an integral of simple convolution in three dimensions.  Thus,
each of these integrals may be computed in $\mathcal{O}(N\log N)$
time, which is the same scaling behavior that is common to the most
common and accurate classical density functional theories, most
notably FMT.

>>>>

\begin{document}
\title{An efficient contact value approach for the pair distribution
  function of the inhomogeneous hard-sphere fluid}

@@

\fixme{
  TO DO:
  \begin{itemize}
  \item \pahosays{Only use short-range CVA-S (in plots, and in walls.cpp)}
  \item \pahosays{Remove CVA from all plots, but leave it easy to
    re-add (so we can check what we say in the text).}
  \item \pahosays{Make all plots not show CVA (new name for CVA-short)
    outside the fitted range.}
  \item \davidsays{Rewrite CVA-S section to reflect new fit.}
  \item \jeffsays{Make certain when proof-reading that other prior
    formulations cannot work in ``separable'' way.}
  \item \davidsays{Be sure that we explain the fitted data and fitting
    process well.}
  \item \pahosays{Add $\eta=?$ labels and legends to new fit plot.}
  \item \jeffsays{I will diligently read what you guys write!}
  \end{itemize}
}

@@

\section{Contact value approach}
In the approaches for the pair distribution function mentioned above,
the radial distribution function used in the approximation
was dependent upon the density averaged over some volume.  We seek to
achieve greater accuracy by making use of a function dependent upon
our averaged $g_{\sigma}(\rr)$ discussed above, which holds more
information about an inhomogeneous system than does a simple
convolution of the density.
%
We construct the CVA with the average of two radial distribution
functions, evaluated at the distance between the two points, that are
themselves functions of the averaged pair distribution function at contact
$g_{\sigma}(\rr)$ evaluated at the two points:
%
%% We therefore seek a function form
%% for our radial distribution that is dependent upon $g_{\sigma}(\rr)$
%% and that will approach the correct solution in the homogeneous limit
%% and also in the limit of contact distance between the two points.
\begin{align}
  g^{(2)}(\rr_1,\rr_2) = \frac{g(r_{12}, g_\sigma(\rr_1)) +
    g(r_{12}, g_\sigma(\rr_2))}{2}. \label{eq:g2-our-mean}
\end{align}
This Contact Value Approach (CVA) for $g^{(2)}(\rr_1,\rr_2)$ is
constructed to reproduce the exact value for the integral:
\begin{align}
  F_1^{\text{contact}} &= \tfrac12 \iint
  g^{(2)}_{HS}(\rr_1,\rr_2)n(\rr_1)n(\rr_2)\delta(|\rr_1-\rr_2|-\sigma)
  d\rr_1d\rr_2
  \label{eq:mean-field-contact}
\end{align}
which is the mean-field correction to the free energy (see
Eq.~\ref{eq:mean-field}) for a purely contact interaction.

@@

\section{Separable radial distribution function for the CVA-S}
\fixme{Section title:  A separable fit for the radial distribution function?}

@@

The key insight of the CVA-S is that we can implement
Eq.~\ref{eq:mean-field} on $\mathcal{O}(N\log N)$ time using Fast
Fourier Transforms (FFTs) where $N$ is proportional to the volume of
the system.  We begin by examining the CVA.  The two terms that are
averaged in Eq.~\ref{eq:g2-our-mean} give equal contributions to the integral:
\begin{align}
  F_1^{\textit{CVA}} &= \tfrac12 \!\! \iint \!\!
  g(r_{12},g_\sigma(\rr_2))n(\rr_1)n(\rr_2)\Phi(|\rr_1-\rr_2|)
  d\rr_1d\rr_2
\end{align}
When we introduce the separable form for $g(r_{12},g_\sigma)$ we can
further simplify this integral as
\begin{multline}
  F_1^{\textit{CVA-S}} =
  \sum_i \tfrac12 \!\! \iint \!\! \bigg( n(\rr_1)  \bigg) \\
                             \bigg( a_i(r_{12})\Phi(r_{12}) \bigg) \\
                             \bigg( b_i(g_\sigma(\rr_2))n(\rr_2) \bigg)
  d\rr_1d\rr_2
\end{multline}
where the functional is written as a summation of integrals, each of
which is an integral of simple convolution in three dimensions.  Thus,
each of these integrals may be computed in $\mathcal{O}(N\log N)$
time, which is the same scaling behavior that is common to the most
common and accurate classical density functional theories, most
notably FMT.
